---
title: "Day 1 Hands-On Exercises: Foundations of R & Tidy Data"
subtitle: "IUCN Bahari Yetu Scholarly Cohort Bootcamp"
format: html
---

## Exercise 1: Directory Setup & the `here` Package

Before running any R code, ensure your workspace is structured logically. You should have:
- A root RStudio Project directory.
- A `data/` subdirectory for raw datasets.
- A `scripts/` or `module/` subdirectory for code.
- An `outputs/` subdirectory for saved figures and tables.

Let's test if the `here` package correctly points to your project root. Run the following code block:

```{r}
#| label: setup-check
# Install if not already installed:
# install.packages("here")

library(here)
print(here())
```

*Question:* What directory does `here()` output? Is it your project's root folder?

---

## Exercise 2: Basic R Operations & Data Structures

### Part A: Vectors
Create a vector representing forest canopy height (in meters) for 5 trees:
`12.4, 15.1, 9.8, 22.3, 14.5`

Create another vector representing the species codes for those 5 trees:
`"T1", "T2", "T1", "T3", "T2"`

```{r}
# Write your code here:
heights <- c(12.4, 15.1, 9.8, 22.3, 14.5)
species <- c("T1", "T2", "T1", "T3", "T2")

# Calculate the mean and standard deviation of tree heights:
mean_height <- mean(heights)
sd_height <- sd(heights)

print(mean_height)
print(sd_height)
```

### Part B: Data Frames
Combine these two vectors into a data frame named `forest_data`. Add a third column for `tree_id` containing the values 1 through 5.

```{r}
# Combine into a data frame
forest_data <- data.frame(
  tree_id = 1:5,
  species = species,
  height = heights
)

print(forest_data)
```

---

## Exercise 3: Importing Data

For this exercise, we will import a mock dataset representing plastic waste counts at different beach sites.

1. Ensure the package `readr` (part of the `tidyverse`) is loaded.
2. Read a CSV file. If you don't have a file, we can write a temporary one to test.

```{r}
library(tidyverse)

# Practice importing the pre-generated plastic waste dataset:
plastics_df <- read_csv(here("module", "data", "theme6_plastic_waste.csv"))
head(plastics_df)
```

---

## Exercise 4: Identifying Untidy Data

Look at the following three tables representing household energy biomass usage.
Which one is **tidy** according to the rules:
1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

### Table A
| Household | Charcoal_Kg_Day1 | Charcoal_Kg_Day2 | Firewood_Kg_Day1 | Firewood_Kg_Day2 |
| :--- | :--- | :--- | :--- | :--- |
| HH01 | 2.5 | 3.0 | 5.0 | 4.5 |
| HH02 | 1.0 | 1.2 | 8.0 | 7.5 |

### Table B
| Household | Fuel_Type | Day | Amount_Kg |
| :--- | :--- | :--- | :--- |
| HH01 | Charcoal | Day 1 | 2.5 |
| HH01 | Charcoal | Day 2 | 3.0 |
| HH01 | Firewood | Day 1 | 5.0 |
| HH01 | Firewood | Day 2 | 4.5 |
| HH02 | Charcoal | Day 1 | 1.0 |
| HH02 | Charcoal | Day 2 | 1.2 |
| HH02 | Firewood | Day 1 | 8.0 |
| HH02 | Firewood | Day 2 | 7.5 |

*Discuss with your theme table:* Why is **Table B** tidy and **Table A** untidy? How does R benefit from the structure of Table B?
