---
title: "Day 4 Elective Track A: Spatial Data & GIS in R"
subtitle: "IUCN Bahari Yetu Scholarly Cohort Bootcamp"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(sf)
library(terra)
library(here)
```

## Exercise 1: Vector Geometries with `sf`

Let's simulate coordinates for plastic waste cleanup stations along the Dar es Salaam coast.

```{r}
# Load the pre-generated plastic waste station coordinates dataset
cleanup_sites <- read_csv(here("module", "data", "theme6_plastic_waste.csv"))

# Convert tabular data to an sf spatial object (using CRS 4326 - WGS 84 GPS coordinates)
cleanup_sites_sf <- st_as_sf(cleanup_sites, coords = c("lon", "lat"), crs = 4326)

print(cleanup_sites_sf)
```

### Tasks
1. Plot the spatial points using `geom_sf()` inside `ggplot()`.
2. Reproject the spatial coordinates from WGS 84 (CRS 4326) to UTM Zone 37S (CRS 32737), which is standard for Tanzania.

```{r}
# Reproject:
cleanup_sites_utm <- st_transform(cleanup_sites_sf, crs = 32737)

# Plot:
ggplot() +
  geom_sf(data = cleanup_sites_utm, aes(size = macroplastics_count, color = macroplastics_count)) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(
    title = "Plastic Waste Cleanup Locations (UTM 37S)",
    size = "Count",
    color = "Count"
  )
```

---

## Exercise 2: Raster Manipulations with `terra`

Let's create a simulated raster representing elevation (DEM) for a 10x10 km area near a coastal forest reserve.

```{r}
#| label: generate-raster
# Create a blank raster grid
dem_raster <- rast(
  nrows = 100, ncols = 100, 
  xmin = 39.2, xmax = 39.3, 
  ymin = -6.8, ymax = -6.7,
  crs = "EPSG:4326"
)

# Fill with simulated elevation values (gradient + noise)
values(dem_raster) <- rep(seq(0, 300, length.out = 100), each = 100) + rnorm(10000, 0, 5)

print(dem_raster)
```

### Tasks
1. Plot the raster elevation map using `plot()`.
2. Extract the elevation values at the plastic cleanup sites coordinates.

```{r}
# Plot raster:
plot(dem_raster, main = "Coastal Digital Elevation Model (DEM)")

# Extract values at points:
extracted_elevation <- terra::extract(dem_raster, cleanup_sites_sf)

# Combine results back:
cleanup_sites_with_elevation <- cleanup_sites %>%
  mutate(elevation_m = extracted_elevation[, 2])

print(cleanup_sites_with_elevation)
```
