---
title: "Day 4 Elective Track B: Statistical Modeling in R"
subtitle: "IUCN Bahari Yetu Scholarly Cohort Bootcamp"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(testflow)
library(rstatix)
library(here)
```

## Exercise 1: Hypothesis Testing with `testflow` & `rstatix`

Let's simulate a dataset comparing soil organic carbon (SOC) levels between two management schemes: **Community-Based Forest Management (CBFM)** and **Joint Forest Management (JFM)**.

```{r}
# Load the pre-generated AFOLU agriculture dataset
forest_soc <- read_csv(here("module", "data", "theme5_ghg_afolu.csv"))

head(forest_soc)
```

### Tasks
1. Execute a two-sample comparison using `testflow::test_t_two_sample()`.
2. Inspect the assumption outputs (normality tests and variance equality).
3. Report the statistical test output and plot diagnostics.
4. **`rstatix` alternative**: Run a t-test using `rstatix::t_test()`, check normality using `rstatix::shapiro_test()`, and calculate effect size using `rstatix::cohens_d()`.

```{r}
# Execute two-sample test using testflow
soc_comparison <- test_t_two_sample(
  data = forest_soc,
  response = "soil_organic_carbon_pct",
  group = "crop_type"
)

# Output scholarly report
report(soc_comparison)

# Plot diagnostics (Shapiro-Wilk, QQ-plot, etc.)
plot(soc_comparison)

# --- rstatix Alternative ---
# 1. Shapiro-Wilk test for normality by group
forest_soc %>% 
  group_by(crop_type) %>% 
  shapiro_test(soil_organic_carbon_pct)

# 2. Run independent t-test
stat_test <- forest_soc %>% 
  t_test(soil_organic_carbon_pct ~ crop_type, detailed = TRUE)
print(stat_test)

# 3. Calculate effect size (Cohen's d)
effect_sz <- forest_soc %>% 
  cohens_d(soil_organic_carbon_pct ~ crop_type)
print(effect_sz)
```

---

## Exercise 2: Linear Regression Modeling

Let's fit a multiple linear regression model predicting mangrove seedling growth (height in cm) based on soil salinity (ppt) and light intensity (PAR).

```{r}
# Load the pre-generated AFOLU dataset for regression
seedlings_data <- read_csv(here("module", "data", "theme5_ghg_afolu.csv"))

head(seedlings_data)
```

### Tasks
1. Fit a linear model using `lm()` and store it in `model_fit`.
2. Print the model summary coefficients using `summary(model_fit)`.
3. Check the residual diagnostic plots using `plot(model_fit, which = 1:2)`.

```{r}
# Fit model (predicting yield based on fertilizer application and soil carbon):
model_fit <- lm(yield_tonnes_ha ~ fertilizer_nitrogen_kg_ha + soil_organic_carbon_pct, data = seedlings_data)

# Show summary:
summary(model_fit)

# Plot diagnostics:
par(mfrow = c(1, 2))
plot(model_fit, which = 1:2)
par(mfrow = c(1, 1))
```
