---
title: "Day 3 Hands-On Exercises: Data Visualization with `ggplot2`"
subtitle: "IUCN Bahari Yetu Scholarly Cohort Bootcamp"
format: html
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
```

## Exercise 1: Building a Basic Scatter Plot

Let's generate simulated marine biology survey data representing dolphin acoustic detections vs. boat traffic density:

```{r}
# Load the pre-generated marine ecology dataset
marine_survey <- read_csv(here("module", "data", "theme4_marine_ecology.csv"))

head(marine_survey)
```

### Tasks
1. Plot `boat_density_index` on the x-axis and `dolphin_detections` on the y-axis using `geom_point()`.
2. Map `survey_site` to the point color aesthetic.
3. Add a linear trendline using `geom_smooth(method = "lm", se = FALSE)`.

```{r}
# Write your plotting pipeline here:
ggplot(marine_survey, aes(x = boat_density_index, y = dolphin_detections, color = site)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Dolphin Acoustic Detections vs. Boat Traffic",
    x = "Boat Density Index",
    y = "Dolphin Detections / Day",
    color = "Survey Site"
  )
```

---

## Exercise 2: Boxplots & Theme Customization

Let's compare the distribution of tree heights across different species using boxplots.

1. Build a boxplot of `tree_species` vs. `dbh_cm` using the `coastal_forest` sample dataset below.
2. Fill the boxplots using the `"viridis"` color palette.
3. Apply `theme_minimal()` and remove the legend (since species is already on the x-axis).

```{r}
# Load the pre-generated coastal forest dataset
coastal_forest <- read_csv(here("module", "data", "theme1_coastal_forest.csv"))

# Plotting code:
ggplot(coastal_forest, aes(x = tree_species, y = dbh_cm, fill = tree_species)) +
  geom_boxplot() +
  scale_fill_viridis_d(option = "plasma") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Tree DBH Distribution by Species",
    x = "Mangrove Species",
    y = "Diameter at Breast Height (cm)"
  )
```

---

## Exercise 3: Faceting

Faceting splits one plot into multiple panels based on a categorical variable.
Create a faceted plot showing dolphin detections vs. boat density, with a separate panel for each `survey_site`.

```{r}
# Plotting code:
ggplot(marine_survey, aes(x = boat_density_index, y = dolphin_detections)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm") +
  facet_wrap(~site) +
  theme_bw() +
  labs(
    title = "Site-wise Dolphin Detections vs. Boat Density",
    x = "Boat Density Index",
    y = "Dolphin Detections / Day"
  )
```

---

## Exercise 4: Exporting Plots with `ggsave()`

Journals require high-resolution (300+ DPI) figures, often with specific width limits (e.g., 90 mm for single-column, 180 mm for double-column layouts).

1. Store your plot in an object named `my_plot`.
2. Save it as a PDF or high-resolution PNG using `ggsave()`.

```{r}
my_plot <- ggplot(marine_survey, aes(x = boat_density_index, y = dolphin_detections, color = site)) +
  geom_point(size = 2.5) +
  theme_classic() +
  labs(
    x = "Boat Density Index",
    y = "Dolphin Detections / Day",
    color = "Site"
  )

# Create outputs folder if it doesn't exist
if (!dir.exists(here("outputs"))) {
  dir.create(here("outputs"))
}

# Export figure:
ggsave(
  filename = here("outputs", "dolphin_detections_plot.png"),
  plot = my_plot,
  width = 150, 
  height = 100, 
  units = "mm", 
  dpi = 300
)
```
