---
title: "Day 1 Session 2 Exercise: Basic R Operations & Data Structures"
subtitle: "Applied Statistics for Marine & Coastal Analytics"
format: html
---

# Part A: Vector Calculations

Create a numeric vector representing forest canopy height (in meters) for 5 trees:
`12.4, 15.1, 9.8, 22.3, 14.5`

Create a character vector representing the species codes for those 5 trees:
`"T1", "T2", "T1", "T3", "T2"`

```{r}
# 1. Assign heights and species vectors:
heights <- c(12.4, 15.1, 9.8, 22.3, 14.5)
species <- c("T1", "T2", "T1", "T3", "T2")

# 2. Calculate the mean and standard deviation of tree heights:
mean_height <- mean(heights)
sd_height <- sd(heights)

print(mean_height)
print(sd_height)
```

# Part B: Creating Data Frames

Combine these two vectors into a data frame named `forest_data`. Include a third column for `tree_id` containing the values 1 through 5.

```{r}
# 3. Combine vectors into a data frame
forest_data <- data.frame(
  tree_id = 1:5,
  species = species,
  height_m = heights
)

print(forest_data)

# 4. Use the structure inspector to see variables and data classes:
str(forest_data)
```

# Discussion Questions
1. What data class is the `species` column? What data class is the `height_m` column?
2. What happens if you try to compute `mean(species)`? Why does this occur?
