---
title: "Reproducible Scholarly Report Template"
subtitle: "Applied Analysis of [Insert Research Theme / Topic]"
author: "Scholar Name (Affiliation: SUA / UDSM / NM-AIST / SUZA)"
date: today
format:
  html:
    toc: true
    number-sections: true
    theme: cosmo
  docx:
    toc: true
    number-sections: true
  pdf:
    toc: true
    number-sections: true
---

# Introduction

*Instructions: Write a brief background of your research topic. This template demonstrates how to compile text, code, tables, and figures into a single document.*

The **IUCN Bahari Yetu Project** supports capacity building in environmental and climate sciences. This report outlines a reproducible analysis pipeline for assessing research parameters.

## Research Questions
1. Is there a statistically significant difference in our response variable across study locations?
2. How does the predictor variable influence our primary metric?

---

# Methodology & Data Import

We read our raw data into R. The chunk below is configured with `#| echo: true` to display the code, but we set `#| warning: false` and `#| message: false` to keep the compiled report clean.

```{r}
#| label: setup-and-load
#| echo: true
#| warning: false
#| message: false

library(tidyverse)
library(flextable)
library(here)

# Load the pre-generated AFOLU dataset
study_data <- read_csv(here("data", "theme5_ghg_afolu.csv"))

head(study_data)
```

---

# Results

## Descriptive Summary

We compute summaries of our measurements grouped by study site. Table @tbl-summary shows the calculated mean and standard deviations.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics of soil organic carbon and crop yield by crop type."
#| echo: false

summary_table <- study_data %>%
  group_by(crop_type) %>%
  summarize(
    `Mean SOC (%)` = round(mean(soil_organic_carbon_pct), 2),
    `SD SOC` = round(sd(soil_organic_carbon_pct), 2),
    `Mean Yield (t/ha)` = round(mean(yield_tonnes_ha), 2),
    `SD Yield` = round(sd(yield_tonnes_ha), 2)
  )

# Format using flextable
flextable(summary_table) %>%
  theme_vanilla() %>%
  autofit()
```

## Visual Analysis

In Figure @fig-biomass, we plot the distribution of biomass across the two sites.

```{r}
#| label: fig-biomass
#| fig-cap: "Comparison of crop yield levels between Maize and Rice farms."
#| echo: false
#| fig-width: 6
#| fig-height: 4

ggplot(study_data, aes(x = crop_type, y = yield_tonnes_ha, fill = crop_type)) +
  geom_boxplot(alpha = 0.8) +
  scale_fill_manual(values = c("#0A3B5C", "#4A7C59")) +
  theme_classic() +
  labs(
    x = "Crop Type",
    y = "Yield (tonnes/ha)",
    fill = "Crop Type"
  )
```

---

# Discussion & Conclusion

*Instructions: Interpret the results shown in Table 1 and Figure 1. Since this pipeline is fully automated in Quarto, any changes to the raw data files will automatically update all numbers, tables, and figures upon re-rendering.*
