---
title: "Day 5 Session 2 Exercise: Cross-Referencing & Formatting"
subtitle: "Applied Statistics for Marine & Coastal Analytics"
format: html
---

# Introduction

This exercise covers advanced layout adjustments inside Quarto:
1. Control code chunk display configurations.
2. Cross-reference generated tables and figures.
3. Export outputs as Word documents.

# Data Setup

We load our datasets inside a silent chunk. Notice that we set `#| echo: false` and `#| message: false` to keep the compiled report clean.

```{r}
#| label: setup-data
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
library(flextable)
library(here)

study_data <- read_csv(here("data", "theme5_ghg_afolu.csv"))
```

# Tasks

1. **Table Reference**: Look at Table @tbl-crop-summary. Verify that the label and caption are defined inside the code chunk options using `#| label: tbl-crop-summary` and `#| tbl-cap: "Summary metrics."`
2. **Figure Reference**: Reference Figure @fig-yield-distribution in the text panel below.

As shown in @fig-yield-distribution, we compare crop yield distributions between Maize and Rice...

```{r}
#| label: tbl-crop-summary
#| tbl-cap: "Mean and standard deviation of crop yields by type."
#| echo: false

crop_table <- study_data %>%
  group_by(crop_type) %>%
  summarize(
    `Mean Yield (t/ha)` = round(mean(yield_tonnes_ha), 2),
    `SD Yield` = round(sd(yield_tonnes_ha), 2)
  )

flextable(crop_table) %>%
  theme_vanilla() %>%
  autofit()
```

```{r}
#| label: fig-yield-distribution
#| fig-cap: "Boxplot comparing crop yields."
#| echo: false
#| fig-width: 5
#| fig-height: 3.5

ggplot(study_data, aes(x = crop_type, y = yield_tonnes_ha, fill = crop_type)) +
  geom_boxplot() +
  theme_bw()
```
